---
title: "data-cleaning"
output: html_document
---

```{r}
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  matchmaker, # dictionary-based cleaning
  epikit,     # age_categories() function
  tidyverse   # data management and visualization
)
```

```{r}
linelist_raw <- import("/home/jeremy/Work/Data Science/Code in Zip/epirhandbook/linelist_raw.xlsx")
linelist_raw
```

```{r}
skimr::skim(linelist_raw)
```

### Automatic Cleaning

```{r}
names(linelist_raw)
```

```{r}
linelist <- linelist_raw %>%
  janitor::clean_names()

names(linelist)
```

### Manual Name Cleaning

```{r}
linelist <- linelist_raw %>%
  # standardize column name syntax
  janitor::clean_names() %>%
  rename(date_infection       = infection_date,
         date_hospitalisation = hosp_date,
         date_outcome         = date_of_outcome)

names(linelist)
```

```{r}
linelist_raw %>% 
  select(# NEW name             # OLD name
         date_infection       = `infection date`,    # rename and KEEP ONLY these columns
         date_hospitalisation = `hosp date`)
```

### Keep Columns

```{r}
linelist %>%
  select(case_id, date_onset, date_infection, date_hospitalisation,
         date_outcome, everything()) %>%
  names()
```

```{r}
linelist %>%
  select(where(is.numeric)) %>%
  names()
```

```{r}
linelist %>%
  select(contains("date")) %>%
  names()
```

-   The function `matches()` works similarly to `contains()` but can be provided a regular example such as multiple strings separated by OR bars within the parentheses

```{r}
linelist %>%
  select(matches("onset|hosp|fev")) %>%
  names()
```

```{r}
linelist %>%
  select(any_of(c("date_onset", "village_origin", 
                "village_residence", "village_travel"))) %>%
  names()
```

### Remove Columns

```{r}
linelist %>%
  select(-c(date_onset, fever:vomit)) %>% # remove date_onset and columns starting from fever to vomit
  names()
```

```{r}
linelist$x28 <- NULL # deletes column with base R syntax
```

### 

```{r}
linelist <- linelist_raw %>%
    janitor::clean_names() %>%                              # standardize column name syntax 
           # NEW name             # OLD name
    rename(date_infection       = infection_date,           # manually re-name columns
           date_hospitalisation = hosp_date,
           date_outcome         = date_of_outcome) %>% 
    select(-c(row_num, merged_header, x28))                 # remove column

names(linelist)
```

### Deduplication

```{r}
linelist <- linelist %>% distinct()
```

```{r}
linelist <- linelist_raw %>%
  janitor::clean_names() %>%
  rename(date_infection       = infection_date,           # manually re-name columns
         date_hospitalisation = hosp_date,
         date_outcome         = date_of_outcome) %>% 
  select(-c(row_num, merged_header, x28)) %>%             # remove column
  distinct()
```

### Transformation

```{r}
linelist <- linelist %>%
  mutate(bmi = wt_kg / (ht_cm/100)^2)

linelist
```

```{r}
new_col_demo <- linelist %>%                       
  mutate(
    new_var_dup    = case_id,             # new column = duplicate/copy another existing column
    new_var_static = 7,                   # new column = all values the same
    new_var_static = new_var_static + 5,  # you can overwrite a column, and it can be a calculation using other variables
    new_var_paste  = stringr::str_glue("{hospital} on ({date_hospitalisation})") # new column = pasting together values from other columns
  ) %>% 
  select(case_id, hospital, date_hospitalisation, contains("new"))        # show only new columns, for demonstration purposes

new_col_demo
```

```{r}
linelist <- linelist %>% select(-contains("new_var"))
```

### Convert Column Class

```{r}
class(linelist$outcome)
```

```{r}
linelist <- linelist %>% 
  mutate(age = as.numeric(age))
```

### Grouped Data

```{r}
# age normalized to mean of ALL rows
linelist %>% 
  mutate(age_norm = age / mean(age, na.rm = T))

# age normalized to mean of hospital group
linelist %>% 
  group_by(hospital) %>% 
  mutate(age_norm = age / mean(age, na.rm = T))
```

```{r}
# Transform multiple columns
linelist <- linelist %>% 
  mutate(across(.cols = c(temp, ht_cm, wt_kg), .fns = as.character))

linelist
```

```{r}
# To change all columns to character class
linelist <- linelist %>% 
  mutate(across(.cols = everything(), .fns = as.character))
```

```{r}
# To change all columns to character class
linelist <- linelist %>% 
  mutate(across(.cols = contains("date"), .fns = as.character))
```

```{r}
linelist <- linelist %>% 
  mutate(across(.cols = where(is.POSIXct), .fns = as.Date))

linelist
```

#### Coalesce

```{r}
names(linelist_raw)
```

```{r}
linelist <- linelist %>%
  mutate(village = coalesce(village_detection, village_residence))

linelist$village
```

```{r}
cumulative_case_counts <- linelist %>%  # begin with case linelist
  count(date_onset) %>%                 # count of rows per day, as column 'n'   
  mutate(cumulative_cases = cumsum(n))  # new column, of the cumulative sum at each row

head(cumulative_case_counts)
```

```{r}
linelist <- linelist_raw %>%
  janitor::clean_names() %>%
  rename(date_infection       = infection_date,           # manually re-name columns
         date_hospitalisation = hosp_date,
         date_outcome         = date_of_outcome) %>% 
  select(-c(row_num, merged_header, x28)) %>%             # remove column
  distinct() %>%
  mutate(bmi = wt_kg / (ht_cm/100)^2) %>%                 # add new column
  mutate(across(contains("date"), as.Date),               # convert class of columns 
         generation = as.numeric(generation),
         age        = as.numeric(age))
```

### Re-code Values

-   `recode()` is designed to handle factors and charater vectors only.

```{r}
# # fix incorrect values                   # old value       # new value
# linelist <- linelist %>% 
#   mutate(date_onset = recode(date_onset, "2014-14-15" = "2014-04-15"))

linelist <- linelist %>%                 
  mutate(date_onset = case_when(
    date_onset == as.Date("2014-14-15", format = "%Y-%d-%m") ~ as.Date("2014-04-15"),
    TRUE ~ date_onset
  ))
```

```{r}
table(linelist$hospital, useNA = "always")  # print table of all unique values, including missing  
```

```{r}
linelist <- linelist %>% 
  mutate(hospital = recode(hospital,
                           # for reference: OLD = NEW
                           "Mitylira Hopital"  = "Military Hospital",
                           "Mitylira Hospital" = "Military Hospital",
                           "Military Hopital"  = "Military Hospital",
                           "Port Hopital"      = "Port Hospital",
                           "Central Hopital"   = "Central Hospital",
                           "other"             = "Other",
                           "St. Marks Maternity Hopital (SMMH)" = "St. Mark's Maternity Hospital (SMMH)"
                     )
         )
```

```{r}
table(linelist$hospital, useNA = "always")
```

### Replace By Logic

```{r}
# Example: change gender of one specific observation to "Female" 
linelist <- linelist %>% 
  mutate(gender = replace(gender, case_id == "2195", "Female"))
```

```{r}
linelist$gender[linelist$case_id == "2195"] <- "Female"
```

```{r}
linelist <- linelist %>% 
  mutate(source_known = ifelse(!is.na(source), "known", "unknown"))
```

```{r}
# Create a column, date_death column, which equates to NA if patient is not yet dead
linelist <- linelist %>%
  mutate(date_death = if_else(outcome == "Death", date_outcome, NA_Date_))
```

```{r}
# Complex logic
linelist <- linelist %>% 
  mutate(age_years = case_when(
    age_unit == "years"  ~ age,       # if age unit is years, just use the age
    age_unit == "months" ~ age/12,    # if age unit is months, new value is the age divided by 12
    is.na(age_unit)      ~ age)       # if age unit is missing, just assume years
  )                                   # any other circumstance, assign NA (missing)

linelist
```

```{r}
linelist <- linelist %>%
  mutate(case_status = case_when(
    ct_blood < 20  ~ "Confirmed",
    !is.na(source) & fever == "yes" ~ "Suspect",
    TRUE ~ "To investigate"
  ))
```

### Missing Values

```{r}
linelist <- linelist %>% 
  mutate(hospital = replace_na(hospital, "Missing"))

linelist
```

`fct_explicit_na()`

```{r}
# For a factor column
linelist <- linelist %>%
  mutate(hospital = fct_explicit_na(hospital))

linelist$hospital %>% head(10)
```

`na_if()`

```{r}
linelist <- linelist %>%
  mutate(hospital = na_if(hospital, "Missing"))
```

```{r}
linelist <- linelist %>%
  mutate(temp = replace(temp, temp > 40, NA))
```

```{r}
linelist <- linelist %>%
  mutate(date_onset = replace(date_onset, date_onset > as.Date("2001-01-01"), NA))
```

### Cleaning dictionary

```{r}
cleaning_dict <- import("/home/jeremy/Work/Data Science/Code in Zip/epirhandbook-master/inst/extdata/case_linelists/cleaning_dict.csv")
cleaning_dict
```

```{r}
linelist <- linelist %>%     # provide or pipe your dataset
     matchmaker::match_df(
          dictionary = cleaning_dict,  # name of your dictionary
          from = "from",               # column with values to be replaced (default is col 1)
          to = "to",                   # column with final values (default is col 2)
          by = "col"                   # column with column names (default is col 3)
  )
```

```{r}
linelist <- linelist_raw %>%
  janitor::clean_names() %>%
  rename(date_infection       = infection_date,           # manually re-name columns
         date_hospitalisation = hosp_date,
         date_outcome         = date_of_outcome) %>% 
  select(-c(row_num, merged_header, x28)) %>%             # remove column
  distinct() %>%
  mutate(bmi = wt_kg / (ht_cm/100)^2) %>%                 # add new column
  mutate(across(contains("date"), as.Date),               # convert class of columns 
         generation = as.numeric(generation),
         age        = as.numeric(age)) %>%
    # add column: delay to hospitalisation
    mutate(days_onset_hosp = as.numeric(date_hospitalisation - date_onset)) %>% 
    # clean values of hospital column
    mutate(hospital = recode(hospital,
                              # OLD = NEW
                              "Mitylira Hopital"  = "Military Hospital",
                              "Mitylira Hospital" = "Military Hospital",
                              "Military Hopital"  = "Military Hospital",
                              "Port Hopital"      = "Port Hospital",
                              "Central Hopital"   = "Central Hospital",
                              "other"             = "Other",
                              "St. Marks Maternity Hopital (SMMH)" = "St. Mark's Maternity Hospital (SMMH)"
                       )
           ) %>% 
    mutate(hospital = replace_na(hospital, "Missing")) %>% 
    # create age_years column (from age and age_unit)
    mutate(age_years = case_when(
          age_unit == "years" ~ age,
          age_unit == "months" ~ age/12,
          is.na(age_unit) ~ age,
          TRUE ~ NA_real_))
```

## Numeric Categories

```{r}
summary(linelist$age_years, na.rm = T)
```

```{r}
# examine the distribution
hist(linelist$age_years)
```

`age_categories()`

```{r}
linelist <- linelist %>%
  mutate(
    age_cat = age_categories(
      age_years,
      breakers = c(0, 5, 10, 15, 20,
                   30, 40, 50, 60, 70)
    )
  )

table(linelist$age_cat, useNA = "always")
```

```{r}
# Include upper ends for the same categories
linelist <- linelist %>% 
  mutate(
    age_cat = age_categories(
      age_years, 
      breakers = c(0, 6, 11, 16, 21, 31, 41, 51, 61, 71),
      separator = "-"))

# show table
table(linelist$age_cat, useNA = "always")
```

```{r}
# With ceiling set to TRUE
linelist <- linelist %>% 
  mutate(
    age_cat = age_categories(
      age_years, 
      breakers = c(0, 5, 10, 15, 20, 30, 40, 50, 60, 70),
      ceiling = TRUE)) # 70 is ceiling, all above become NA

# show table
table(linelist$age_cat, useNA = "always")
```

```{r}
linelist <- linelist %>% 
  mutate(
    age_cat = age_categories(
      age_years, 
      lower = 0,
      upper = 100,
      by = 10))

# show table
table(linelist$age_cat, useNA = "always")
```

`cut()`

```{r}
# Create new variable, by cutting the numeric age variable lower break is excluded but upper break is included in each category
linelist <- linelist %>% 
  mutate(
    age_cat = cut(
      age_years,
      breaks = c(0, 5, 10, 15, 20,
                 30, 50, 70, 100),
      include.lowest = TRUE         # include 0 in lowest group
      ))

# tabulate the number of observations per group
table(linelist$age_cat, useNA = "always")
```

```{r}
# Cross tabulation of the numeric and category columns. 
table("Numeric Values" = linelist$age_years,   # names specified in table for clarity.
      "Categories"     = linelist$age_cat,
      useNA = "always")                        # don't forget to examine NA values
```

```{r}
linelist <- linelist %>% 
  # cut() creates age_cat, automatically of class Factor      
  mutate(
    age_cat = cut(
      age_years,
      breaks = c(0, 5, 10, 15, 20, 30, 50, 70, 100),          
      right = FALSE,
      include.lowest = TRUE,        
      labels = c("0-4", "5-9", "10-14", "15-19", "20-29", "30-49", "50-69", "70-100")
    ),
         
    # make missing values explicit
    age_cat = fct_explicit_na(
      age_cat,
      na_level = "Missing age"
    )  # you can specify the label
  )

linelist
```

```{r}
# table to view counts
table(linelist$age_cat, useNA = "always")
```

### Quantile Breaks

```{r}
quantile(
  linelist$age_years,               # specify numeric vector to work on
  probs = c(0, .25, .50, .75, .90, .95),   # specify the percentiles you want
  na.rm = TRUE
)                            # ignore missing values 
```

```{r}
linelist %>%                                # begin with linelist
  mutate(deciles = cut(age_years,           # create new column decile as cut() on column age_years
    breaks = quantile(                      # define cut breaks using quantile()
      age_years,                               # operate on age_years
      probs = seq(0, 1, by = 0.1),             # 0.0 to 1.0 by 0.1
      na.rm = TRUE),                           # ignore missing values
    include.lowest = TRUE)) %>%             # for cut() include age 0
  janitor::tabyl(deciles)                   # pipe to table to display
```

### Evenly Size Groups

```{r}
# make groups with ntile()
ntile_data <- linelist %>% 
  mutate(even_groups = ntile(age_years, 10))

# make table of counts and proportions by group
ntile_table <- ntile_data %>% 
  janitor::tabyl(even_groups)
  
# attach min/max values to demonstrate ranges
ntile_ranges <- ntile_data %>% 
  group_by(even_groups) %>% 
  summarise(
    min = min(age_years, na.rm=T),
    max = max(age_years, na.rm=T)
  )
```

```{r}
# combine and print - note that values are present in multiple groups
left_join(ntile_table, ntile_ranges, by = "even_groups")
```

```{r}
linelist <- linelist_raw %>%
  janitor::clean_names() %>%
  rename(date_infection       = infection_date,           # manually re-name columns
         date_hospitalisation = hosp_date,
         date_outcome         = date_of_outcome) %>% 
  select(-c(row_num, merged_header, x28)) %>%             # remove column
  distinct() %>%
  mutate(bmi = wt_kg / (ht_cm/100)^2) %>%                 # add new column
  mutate(across(contains("date"), as.Date),               # convert class of columns 
         generation = as.numeric(generation),
         age        = as.numeric(age)) %>%
  # add column: delay to hospitalisation
  mutate(days_onset_hosp = as.numeric(date_hospitalisation - date_onset)) %>% 
  # clean values of hospital column
  mutate(hospital = recode(hospital,
                            # OLD = NEW
                            "Mitylira Hopital"  = "Military Hospital",
                            "Mitylira Hospital" = "Military Hospital",
                            "Military Hopital"  = "Military Hospital",
                            "Port Hopital"      = "Port Hospital",
                            "Central Hopital"   = "Central Hospital",
                            "other"             = "Other",
                            "St. Marks Maternity Hopital (SMMH)" = "St. Mark's Maternity Hospital (SMMH)"
                     )
         ) %>% 
  mutate(hospital = replace_na(hospital, "Missing")) %>% 
  # create age_years column (from age and age_unit)
  mutate(age_years = case_when(
        age_unit == "years" ~ age,
        age_unit == "months" ~ age/12,
        is.na(age_unit) ~ age,
        TRUE ~ NA_real_)) %>% 
  mutate(
    # age categories: custom
    age_cat = epikit::age_categories(age_years, breakers = c(0, 5, 10, 15, 20, 30, 50, 70)),
    # age categories: 0 to 85 by 5s
    age_cat5 = epikit::age_categories(age_years, breakers = seq(0, 85, 5))
  )
```

### Add Rows

```{r}
linelist <- linelist %>% 
  add_row(row_num = 666,
          case_id = "abc",
          generation = 4,
          date_infection = as.Date("2020-10-10"),
          .before = 2)
```

### Filter Rows

```{r}
# View first 100 rows
linelist %>% head(100)     # or use tail() to see the n last rows

# Show row 5 only
linelist %>% filter(row_number() == 5)

# View rows 2 through 20, and three specific columns
linelist %>% filter(row_number() %in% 2:20) %>% select(date_onset, outcome, age)
```

```{r}
linelist <- linelist %>% 
  # keep rows where onset is after 1 June 2013 OR where onset is missing and it was a hospital OTHER than Hospital A or B
  filter(date_onset > as.Date("2013-06-01") | (is.na(date_onset) & !hospital %in% c("Hospital A", "Hospital B")))

nrow(linelist)
```

```{r}
table(Hospital  = linelist$hospital,                     # hospital name
      YearOnset = lubridate::year(linelist$date_onset),  # year of date_onset
      useNA     = "always")                              # show missing values
```

```{r}
# dataframe <- filter(dataframe, condition(s) for rows to keep)
linelist <- filter(linelist, !is.na(case_id))
linelist
```

```{r}
linelist <- linelist_raw %>%
  janitor::clean_names() %>%
  rename(date_infection       = infection_date,           # manually re-name columns
         date_hospitalisation = hosp_date,
         date_outcome         = date_of_outcome) %>% 
  select(-c(row_num, merged_header, x28)) %>%             # remove column
  distinct() %>%
  mutate(bmi = wt_kg / (ht_cm/100)^2) %>%                 # add new column
  mutate(across(contains("date"), as.Date),               # convert class of columns 
         generation = as.numeric(generation),
         age        = as.numeric(age)) %>%
  # add column: delay to hospitalisation
  mutate(days_onset_hosp = as.numeric(date_hospitalisation - date_onset)) %>% 
  # clean values of hospital column
  mutate(hospital = recode(hospital,
                            # OLD = NEW
                            "Mitylira Hopital"  = "Military Hospital",
                            "Mitylira Hospital" = "Military Hospital",
                            "Military Hopital"  = "Military Hospital",
                            "Port Hopital"      = "Port Hospital",
                            "Central Hopital"   = "Central Hospital",
                            "other"             = "Other",
                            "St. Marks Maternity Hopital (SMMH)" = "St. Mark's Maternity Hospital (SMMH)"
                     )
         ) %>% 
  mutate(hospital = replace_na(hospital, "Missing")) %>% 
  # create age_years column (from age and age_unit)
  mutate(age_years = case_when(
        age_unit == "years" ~ age,
        age_unit == "months" ~ age/12,
        is.na(age_unit) ~ age,
        TRUE ~ NA_real_)) %>% 
  mutate(
    # age categories: custom
    age_cat = epikit::age_categories(age_years, breakers = c(0, 5, 10, 15, 20, 30, 50, 70)),
    # age categories: 0 to 85 by 5s
    age_cat5 = epikit::age_categories(age_years, breakers = seq(0, 85, 5))
  ) %>%
  filter(
          # keep only rows where case_id is not missing
          !is.na(case_id),  
          
          # also filter to keep only the second outbreak
          date_onset > as.Date("2013-06-01") | (is.na(date_onset) & !hospital %in% 
                                                  c("Hospital A", "Hospital B")))
```

### Row-wise calculations

```{r}
linelist %>%
  rowwise() %>%
  mutate(num_symptoms = sum(c(fever, chills, cough, aches, vomit) == "yes")) %>% 
  ungroup() %>% 
  select(fever, chills, cough, aches, vomit, num_symptoms) # for display
```

```{r}
linelist %>%
  rowwise() %>%
  mutate(num_NA_dates = sum(is.na(c_across(contains("date"))))) %>% 
  ungroup() %>% 
  select(num_NA_dates, contains("date")) # for display
```

```{r}
linelist %>%
  rowwise() %>%
  mutate(latest_date = max(c_across(contains("date")), na.rm=T)) %>% 
  ungroup() %>% 
  select(latest_date, contains("date"))  # for display
```

### Arrange and Sort

```{r}
linelist %>% 
   arrange(hospital, desc(date_onset))
```
