---
title: "Factors"
output: html_document
date: "2025-05-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
pacman::p_load(
  rio,           # import/export
  here,          # filepaths
  lubridate,     # working with dates
  forcats,       # factors
  aweek,         # create epiweeks with automatic factor levels
  janitor,       # tables
  tidyverse      # data mgmt and viz
)
```

```{r}
linelist_cleaned <- import("/home/jeremy/Work/Development/Projects/epidemic/datasets/linelist_cleaned.rds")
head(linelist_cleaned)
```

```{r}
linelist <- linelist_cleaned
```

```{r}
linelist <- linelist %>% 
  mutate(delay_cat = case_when(
    # criteria                                   # new value if TRUE
    days_onset_hosp < 2                        ~ "<2 days",
    days_onset_hosp >= 2 & days_onset_hosp < 5 ~ "2-5 days",
    days_onset_hosp >= 5                       ~ ">5 days",
    is.na(days_onset_hosp)                     ~ NA_character_,
    TRUE                                       ~ "Check me"))
```

```{r}
table(linelist$delay_cat, useNA = "always")
```

```{r}
ggplot(data = linelist)+
  geom_bar(mapping = aes(x = delay_cat))
```

## Convert to Factor

```{r}
linelist <- linelist %>%
  mutate(delay_cat = fct_relevel(delay_cat))

levels(linelist$delay_cat)
```

The function `fct_relevel()` has the additional utility of allowing you to manually specify the level order. Simply write the level values in order, in quotation marks, separated by commas, as shown below. Note that the spelling must exactly match the values. If you want to create levels that do not exist in the data, use [`fct_expand()` instead](https://www.epirhandbook.com/en/new_pages/factors.html#fct_add)).

```{r}
linelist <- linelist %>%
  mutate(delay_cat = fct_relevel(delay_cat, "<2 days", "2-5 days", ">5 days"))

levels(linelist$delay_cat)
```

```{r}
ggplot(data = linelist)+
  geom_bar(mapping = aes(x = delay_cat))
```

## Add or Drop Levels

If you need to add levels to a factor, you can do this with `fct_expand()`. Just write the column name followed by the new levels (separated by commas). By tabulating the values, we can see the new levels and the zero counts. You can use `table()` from **base** R, or `tabyl()` from **janitor**:

```{r}
# Add
linelist %>% 
  mutate(delay_cat = fct_expand(delay_cat, "Not admitted to hospital", "Transfer to other jurisdiction")) %>% 
  tabyl(delay_cat)   # print table
```

If you use `fct_drop()`, the “unused” levels with zero counts will be dropped from the set of levels. The levels we added above (“Not admitted to a hospital”) exists as a level but no rows actually have those values. So they will be dropped by applying `fct_drop()` to our factor column:

```{r}
# Drop
linelist %>% 
  mutate(delay_cat = fct_drop(delay_cat)) %>% 
  tabyl(delay_cat)
```

```{r}
# re-define level order
linelist <- linelist %>% 
  mutate(delay_cat = fct_relevel(delay_cat, c("<2 days", "2-5 days", ">5 days")))
```

```{r}
# re-define level order
linelist %>% 
  mutate(delay_cat = fct_relevel(delay_cat, "<2 days", after = 1)) %>% 
  tabyl(delay_cat)
```

```{r}
# Alpha-numeric default order - no adjustment within ggplot
ggplot(data = linelist)+
    geom_bar(mapping = aes(x = delay_cat))

# Factor level order adjusted within ggplot
ggplot(data = linelist)+
  geom_bar(mapping = aes(x = fct_relevel(delay_cat, c("<2 days", "2-5 days", ">5 days"))))
```

```{r}
# By frequency
# ordered by frequency
ggplot(data = linelist, aes(x = fct_infreq(delay_cat)))+
  geom_bar()+
  labs(x = "Delay onset to admission (days)",
       title = "Ordered by frequency")

# reversed frequency
ggplot(data = linelist, aes(x = fct_rev(fct_infreq(delay_cat))))+
  geom_bar()+
  labs(x = "Delay onset to admission (days)",
       title = "Reverse of order by frequency")
```

```{r}
# By Appearance
# boxplots ordered by original factor levels
ggplot(data = linelist)+
  geom_boxplot(
    aes(x = delay_cat,
        y = ct_blood, 
        fill = delay_cat))+
  labs(x = "Delay onset to admission (days)",
       title = "Ordered by original alpha-numeric levels")+
  theme_classic()+
  theme(legend.position = "none")


# boxplots ordered by median CT value
ggplot(data = linelist)+
  geom_boxplot(
    aes(x = fct_reorder(delay_cat, ct_blood, "median"),
        y = ct_blood,
        fill = delay_cat))+
  labs(x = "Delay onset to admission (days)",
       title = "Ordered by median CT value in group")+
  theme_classic()+
  theme(legend.position = "none")
```

```{r}
# By End Value
epidemic_data <- linelist %>%         # begin with the linelist   
    filter(date_onset < as.Date("2014-09-21")) %>%    # cut-off date, for visual clarity
    count(                                            # get case counts per week and by hospital
      epiweek = lubridate::floor_date(date_onset, "week"),  
      hospital                                            
    ) 
  
ggplot(data = epidemic_data)+                       # start plot
  geom_line(                                        # make lines
    aes(
      x = epiweek,                                  # x-axis epiweek
      y = n,                                        # height is number of cases per week
      color = fct_reorder2(hospital, epiweek, n)))+ # data grouped and colored by hospital, with factor order by height at end of plot
  labs(title = "Factor levels (and legend display) by line height at end of plot",
       color = "Hospital")                          # change legend title
```

## Missing Values

```{r}
linelist %>% 
  mutate(delay_cat = fct_explicit_na(delay_cat, na_level = "Missing delay")) %>% 
  tabyl(delay_cat)
```

```{r}
levels(linelist$delay_cat)
```

The new levels are created using syntax `fct_recode(column, "new" = "old", "new" = "old", "new" = "old")` and printed:

```{r}
linelist %>% 
  mutate(delay_cat = fct_recode(
    delay_cat,
    "Less than 2 days" = "<2 days",
    "2 to 5 days"      = "2-5 days",
    "More than 5 days" = ">5 days")) %>% 
  tabyl(delay_cat)
```

## Reduce into 'Other'

You can use `fct_other()` to manually assign factor levels to an “Other” level. Below, all levels in the column `hospital`, aside from “Port Hospital” and “Central Hospital”, are combined into “Other”. You can provide a vector to either `keep =`, or `drop =`. You can change the display of the “Other” level with `other_level =`.

```{r}
linelist %>%    
  mutate(hospital = fct_other(                      # adjust levels
    hospital,
    keep = c("Port Hospital", "Central Hospital"),  # keep these separate
    other_level = "Other Hospital")) %>%            # All others as "Other Hospital"
  tabyl(hospital)                                   # print table
```

## Reduce into Frequency

```{r}
linelist %>%    
  mutate(hospital = fct_lump(                      # adjust levels
    hospital,
    n = 2,                                          # keep top 2 levels
    other_level = "Other Hospital")) %>%            # all others as "Other Hospital"
  tabyl(hospital)                                   # print table
```

```{r}
ggplot(data = linelist)+
  geom_bar(mapping = aes(x = hospital, fill = age_cat)) +
  scale_fill_discrete(drop = FALSE)+                        # show all age groups in the legend, even those not present
  labs(
    title = "All age groups will appear in legend, even if not present in data")
```

```{r}
linelist %>% 
  mutate(epiweek_date = floor_date(date_onset, "week")) %>%  # create week column
  ggplot()+                                                  # begin ggplot
  geom_histogram(mapping = aes(x = epiweek_date))+           # histogram of date of onset
  scale_x_date(date_labels = "%Y-W%W")                       # adjust disply of dates to be YYYY-WWw
```

```{r}
linelist <- linelist %>% 
  mutate(epiweek_date = floor_date(date_onset, "week"),       # create epiweeks (YYYY-MM-DD)
         epiweek_formatted = format(epiweek_date, "%Y-W%W"),  # Convert to display (YYYY-WWw)
         epiweek_formatted = factor(epiweek_formatted))       # Convert to factor

# Display levels
levels(linelist$epiweek_formatted)
```

```{r}
df <- linelist %>% 
  mutate(epiweek = date2week(date_onset, week_start = "Monday", factor = TRUE))

levels(df$epiweek)
```
