---
title: "Grouping Data"
output: pdf
date: "2025-05-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pacman::p_load(
  rio,       # to import data
  here,      # to locate files
  tidyverse, # to clean, handle, and plot the data (includes dplyr)
  janitor)   # adding total rows and columns
```

```{r}
linelist <- import("/home/jeremy/Work/Development/Projects/epidemic/datasets/linelist_cleaned.rds")
linelist
```

```{r}
ll_by_outcome <- linelist %>% group_by(outcome)
ll_by_outcome
```

## Grouping

```{r}
linelist %>% group_by(outcome) %>% tally()
```

```{r}
linelist %>% 
  group_by(outcome, gender) %>% 
  tally()
```

```{r}
# group dat based on a binary column created *within* the group_by() command
linelist %>% 
  group_by(age_class = ifelse(age >= 18, "adult", "child")) %>% 
  tally(sort = T)
```

By default, if you run `group_by()` on data that are already grouped, the old groups will be removed and the new one(s) will apply. If you want to add new groups to the existing ones, include the argument `.add = TRUE`.

```{r}
# Grouped by outcome
by_outcome <- linelist %>% 
  group_by(outcome)

# Add grouping by gender in addition
by_outcome_gender <- by_outcome %>% 
  group_by(gender, .add = TRUE)
```

If you group on a column of class factor there may be levels of the factor that are not currently present in the data. If you group on this column, by default those non-present levels are dropped and not included as groups. To change this so that all levels appear as groups (even if not present in the data), set `.drop = FALSE` in your `group_by()` command.

## Un-group

```{r}
linelist %>% 
  group_by(outcome, gender) %>% 
  tally() %>% 
  ungroup()
```

You can also remove grouping for only specific columns, by placing the column name inside `ungroup()`.

```{r}
linelist %>% 
  group_by(outcome, gender) %>% 
  tally() %>% 
  ungroup(gender) # remove the grouping by gender, leave grouping by outcome
```

## Summarize

```{r}
# summary statistics on ungrouped linelist
linelist %>% 
  summarise(
    n_cases  = n(),
    mean_age = mean(age_years, na.rm=T),
    max_age  = max(age_years, na.rm=T),
    min_age  = min(age_years, na.rm=T),
    n_males  = sum(gender == "m", na.rm=T))
```

```{r}
# summary statistics on grouped linelist
linelist %>% 
  group_by(outcome) %>% 
  summarise(
    n_cases  = n(),
    mean_age = mean(age_years, na.rm=T),
    max_age  = max(age_years, na.rm=T),
    min_age  = min(age_years, na.rm=T),
    n_males    = sum(gender == "m", na.rm=T))
```

## Counts and Tallies

```{r}
linelist %>% 
  tally()
```

```{r}
linelist %>% 
  group_by(outcome) %>% 
  tally(sort = TRUE)
```

```{r}
linelist %>% 
  count(outcome)
```

```{r}
linelist %>% 
  count(age_class = ifelse(age >= 18, "adult", "child"), sort = T)
```

```{r}
linelist %>% 
  # produce counts by unique outcome-gender groups
  count(gender, hospital) %>% 
  # gather rows by gender (3) and count number of hospitals per gender (6)
  count(gender, name = "hospitals per gender" ) 
```

In contrast to `count()` and `summarise()`, you can use `add_count()` to *add* a new column `n` with the counts of rows per group *while retaining all the other data frame columns*.

```{r}
linelist %>% 
  as_tibble() %>%                   # convert to tibble for nicer printing 
  add_count(hospital) %>%           # add column n with counts by hospital
  select(hospital, n, everything()) # re-arrange for demo purposes
```

```{r}
linelist %>%                                  # case linelist
  tabyl(age_cat, gender) %>%                  # cross-tabulate counts of two columns
  adorn_totals(where = "row") %>%             # add a total row
  adorn_percentages(denominator = "col") %>%  # convert to proportions with column denominator
  adorn_pct_formatting() %>%                  # convert proportions to percents
  adorn_ns(position = "front") %>%            # display as: "count (percent)"
  adorn_title(                                # adjust titles
    row_name = "Age Category",
    col_name = "Gender")
```

## Grouping by date

```{r}
daily_counts <- linelist %>% 
  drop_na(date_onset) %>%        # remove that were missing date_onset
  count(date_onset)              # count number of rows per unique date

daily_counts
```
