---
title: "Characters and Strings"
output: html_document
date: "2025-05-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pacman::p_load(
  stringr,    # many functions for handling strings
  tidyverse,  # for optional data manipulation
  tools       # alternative for converting title cases
)
```

```{r}
linelist_cleaned <- import("/home/jeremy/Work/Development/Projects/epidemic/datasets/linelist_cleaned.rds")
head(linelist_cleaned)
```

```{r}
# Combine Strings
str_c("String1", "String2", "String3")
```

```{r}
str_c("String1", "String2", "String3")
```

```{r}
str_c("String1", "String2", "String3", sep = ", ")
```

```{r}
first_names <- c("abdul", "fahruk", "janice") 
last_names  <- c("hussein", "akinleye", "okeke")

# sep displays between the respective input strings, while collapse displays between the elements produced
str_c(first_names, last_names, sep = " ", collapse = ";  ")
```

```{r}
# For newlines to print correctly, the phrase may need to be wrapped in cat()
cat(str_c(first_names, last_names, sep = " ", collapse = ";\n"))
```

```{r}
# Dynamic Strings
str_glue("Data include {nrow(linelist_cleaned)} cases and are current to {format(Sys.Date(), '%d %b %Y')}.")
```

```{r}
str_glue("Linelist as of {current_date}.\nLast case hospitalized on {last_hospital}.\n{n_missing_onset} cases are missing date of onset and not shown",
         current_date = format(Sys.Date(), '%d %b %Y'),
         last_hospital = format(as.Date(max(linelist$date_hospitalisation, na.rm=T)), '%d %b %Y'),
         n_missing_onset = nrow(linelist %>% filter(is.na(date_onset)))
         )
```

Pulling from a dataframe

```{r}
# make case data frame
case_table <- data.frame(
  zone        = c("Zone 1", "Zone 2", "Zone 3", "Zone 4", "Zone 5"),
  new_cases   = c(3, 0, 7, 0, 15),
  total_cases = c(40, 4, 25, 10, 103)
  )
```

```{r}
case_table %>% 
  str_glue_data("{zone}: {new_cases} ({total_cases} total cases)")
```

You can make the statement appear in one line using str_c() (specifying the data frame and column names), and providing sep = and collapse = arguments.

```{r}
str_c(case_table$zone, case_table$new_cases, sep = " = ", collapse = ";  ")
```

```{r}
str_c("New Cases: ", str_c(case_table$zone, case_table$new_cases, sep = " = ", collapse = ";  "))
```

## UNITE COLUMNS

```{r}
df <- data.frame(
  case_ID = c(1:6),
  symptoms  = c("jaundice, fever, chills",     # patient 1
                "chills, aches, pains",        # patient 2 
                "fever",                       # patient 3
                "vomiting, diarrhoea",         # patient 4
                "bleeding from gums, fever",   # patient 5
                "rapid pulse, headache"),      # patient 6
  outcome = c("Recover", "Death", "Death", "Recover", "Recover", "Recover"))

df_split <- separate(df, symptoms, into = c("sym_1", "sym_2", "sym_3"), extra = "merge")
df_split
```

Then unite the three symptoms column

```{r}
df_split %>% 
  unite(
    col = "all_symptoms",         # name of the new united column
    c("sym_1", "sym_2", "sym_3"), # columns to unite
    sep = ", ",                   # separator to use in united column
    remove = TRUE,                # if TRUE, removes input cols from the data frame
    na.rm = TRUE                  # if TRUE, missing values are removed before uniting
  )
```

```{r}
str_split(string = "jaundice, fever, chills",
          pattern = ",")
```

```{r}
symptoms <- c("jaundice, fever, chills",     # patient 1
              "chills, aches, pains",        # patient 2 
              "fever",                       # patient 3
              "vomiting, diarrhoea",         # patient 4
              "bleeding from gums, fever",   # patient 5
              "rapid pulse, headache")       # patient 6

str_split(symptoms, ",")                     # split each patient's symptoms
```

```{r}
str_split(symptoms, ",", simplify = TRUE)
```

```{r}
str_split(symptoms, ",", simplify = TRUE, n = 2)
```

str_split_fixed(), in which you do not give the simplify argument, but must instead designate the number of columns (n).

```{r}
str_split_fixed(symptoms, ",", n = 2)
```

Split Columns

```{r}
# third symptoms combined into second new column
df %>% 
  separate(symptoms, into = c("sym_1", "sym_2"), sep=",", extra = "merge")
```

```{r}
# third symptoms combined into second new column
df %>% 
  separate(symptoms, into = c("sym_1", "sym_2"), sep=",", extra = "drop")
```

## ARRANGE ALPHABETICALLY

```{r}
# strings
health_zones <- c("Alba", "Takota", "Delta")

# return the alphabetical order
str_order(health_zones)
```

```{r}
str_sort(health_zones)
```

```{r}
n_beds <- 10
n_masks <- 20

paste0("Regional hospital needs ", n_beds, " beds and ", n_masks, " masks.")
```

```{r}
str_to_upper("California")
str_to_lower("California")
```

```{r}
str_to_title("go to the US state of california ")
```

```{r}
tools::toTitleCase("This is the US state of california")
```

```{r}
str_to_sentence("the patient must be transported")
```

## PAD LENGTH
Use str_pad() to add characters to a string, to a minimum length. By default spaces are added, but you can also pad with other characters using the pad = argument.

```{r}
# ICD codes of differing length
ICD_codes <- c("R10.13",
               "R10.819",
               "R17")

# ICD codes padded to 7 characters on the right side
str_pad(ICD_codes, 7, "right")

# Pad with periods instead of spaces
str_pad(ICD_codes, 7, "right", pad = ".")

# Add leading zeros to two digits (e.g. for times minutes/hours)
str_pad("4", 2, pad = "0") 
```

## TRUNCATION

```{r}
original <- "Symptom onset on 4/3/2020 with vomiting"
str_trunc(original, 10, "center")
```

## STANDARDIZE LENGTH

```{r}
# ICD codes of differing length
ICD_codes   <- c("R10.13",
                 "R10.819",
                 "R17")

# truncate to maximum length of 6
ICD_codes_2 <- str_trunc(ICD_codes, 6)
ICD_codes_2
```

```{r}
# expand to minimum length of 6
ICD_codes_3 <- str_pad(ICD_codes_2, 6, "right")
ICD_codes_3
```

## REMOVING TRAILING/LEADING WHITESPACES

```{r}
# ID numbers with excess spaces on right
IDs <- c("provA_1852  ", # two excess spaces
         "provA_2345",   # zero excess spaces
         "provA_9460 ")  # one excess space

# IDs trimmed to remove excess spaces on right side only
str_trim(IDs)
```

```{r}
# original contains excess spaces within string
str_squish("  Pt requires   IV saline\n") 
```

```{r}
pt_course <- "Symptom onset 1/4/2020 vomiting chills fever. Pt saw traditional healer in home village on 2/4/2020. On 5/4/2020 pt symptoms worsened and was admitted to Lumta clinic. Sample was taken and pt was transported to regional hospital on 6/4/2020. Pt died at regional hospital on 7/4/2020."

str_wrap(pt_course, 40)
```

```{r}
cat(str_wrap(pt_course, 40))
```

```{r}
# start and end third from left (3rd letter from left)
str_sub("pneumonia", 3, 3)
```

```{r}
# 6th from left, to the 1st from right
str_sub("pneumonia", 6, -1)
# 6th from left, to the 1st from right
str_sub("pneumonia", -4, -2)
# 4th from left to a position outside the string
str_sub("pneumonia", 4, 15)
```

## Extract by word position

```{r}
# strings to evaluate
chief_complaints <- c("I just got out of the hospital 2 days ago, but still can barely breathe.",
                      "My stomach hurts",
                      "Severe ear pain")

# extract 1st to 3rd words of each string
word(chief_complaints, start = 1, end = 3, sep = " ")
```

```{r}
word <- "pneumonia"

# convert the third and fourth characters to X 
str_sub(word, 3, 4) <- "XX"

# print
word
```

```{r}
words <- c("pneumonia", "tubercolosis", "HIV")

# convert the third and fourth characters to X 
str_sub(words, 3, 4) <- "XX"

words
```

```{r}
str_length("abc")
```

## Detect Patterns

```{r}
str_detect(string = "primary school teacher", pattern = "teach")

# The argument negate = can be included and set to TRUE if you want to know if the pattern is NOT present.
str_detect(string = "primary school teacher", pattern = "teach", negate = TRUE)

# To ignore case/capitalization, wrap the pattern within regex(), and within 
# regex() add the argument ignore_case = TRUE (or T as shorthand).
str_detect(string = "Teacher", pattern = regex("teach", ignore_case = T))
```

```{r}
# a vector/column of occupations 
occupations <- c("field laborer",
                 "university professor",
                 "primary school teacher & tutor",
                 "tutor",
                 "nurse at regional hospital",
                 "lineworker at Amberdeen Fish Factory",
                 "physican",
                 "cardiologist",
                 "office worker",
                 "food service")

# Detect presence of pattern "teach" in each string - output is vector of TRUE/FALSE
str_detect(occupations, "teach")
```

```{r}
# If you need to count the TRUEs, simply sum() the output. This counts the number TRUE.
sum(str_detect(occupations, "teach"))
```

```{r}
# To search inclusive of multiple terms, include them separated by OR bars (|) 
# within the pattern = argument, as shown below:
sum(str_detect(string = occupations, pattern = "teach|professor|tutor"))
```

If you need to build a long list of search terms, you can combine them using str_c() and sep = |, then define this is a character object, and then reference the vector later more succinctly. The example below includes possible occupation search terms for front-line medical providers.

```{r}
# search terms
occupation_med_frontline <- str_c("medical", "medicine", "hcw", "healthcare", "home care", "home health",
                                "surgeon", "doctor", "doc", "physician", "surgery", "peds", "pediatrician",
                               "intensivist", "cardiologist", "coroner", "nurse", "nursing", "rn", "lpn",
                               "cna", "pa", "physician assistant", "mental health",
                               "emergency department technician", "resp therapist", "respiratory",
                                "phlebotomist", "pharmacy", "pharmacist", "hospital", "snf", "rehabilitation",
                               "rehab", "activity", "elderly", "subacute", "sub acute",
                                "clinic", "post acute", "therapist", "extended care",
                                "dental", "dential", "dentist", sep = "|")

occupation_med_frontline
```

```{r}
sum(str_detect(string = occupations, pattern = occupation_med_frontline))
```

```{r}
outcome <- c("Karl: dead",
            "Samantha: dead",
            "Marco: not dead")

str_replace_all(string = outcome, pattern = "dead", replacement = "deceased")
```

